---
title: "CRP-analysis"
author: "GH"
date: "26 2 2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(regclass)  # if not available, install with install.packages("regclass")
```



# Analysis of blood analysis data set

## Read data set

```{r}
tab1<-read.csv2("data_blood.csv")
attach(tab1)
```

## Sum of five subtypes equals white blood cell counts

```{r}
sum_cells <- NEU + EOS + BASO + LYM + MONO

plot(sum_cells, WBC)
abline(0,1, lty=3)
```

## Pairwise correlations

```{r}
round(cor(cbind(NEU, EOS, BASO, LYM, MONO, WBC)),3)
```
## Exact collineartiy

The analysis of the basic regression model will result in the program setting one of the coefficients to NA.

```{r}
mod <- lm(CRP ~ NEU + EOS + BASO + LYM + MONO + WBC)
summary(mod)
```

Next, VIFs are computed using the package `regclass`. The function `VIF` results in an error as there is exact collinearity.

```{r}
try(VIF(mod))
```
## Near collinearity

Now we consider a model with just two variables, `NEU` and `WBC`. 
First we investigate correlation.

```{r}
plot(NEU, WBC)
abline(0,1,lty=3)
text(0,20, paste("Correlation = ", round(cor(NEU, WBC),3)), adj=0)
legend("bottomrigh", lty=3, legend="Diagonal")
```

Now we fit the model and evaluate the VIFs.

```{r}
mod2 <- lm(CRP ~ NEU + WBC)
VIF(mod2)
```
Evaluate the model:

```{r}
summary(mod2)
```

The regression coefficients are quite extreme, but this results from their meaning:

The regression coefficient of `NEU` has the interpretation of the expected difference in `CRP` corresponding to a difference of 1 G/L in `NEU` but keeping `WBC` constant. From the plot above we see that it is sometimes not possible to increase `NEU` by 1 G/L and at the same time holding `WBC` fixed, as `NEU` is a component of `WBC` and cannot exceed `WBC`. Hence the regression coefficient has a problematic interpretation.

Remedy: compute a new variable `WBCnoNEU = WBC - NEU`

```{r}
WBCnoNEU <- WBC - NEU
```

Evaluate correlation again:

```{r}
plot(NEU, WBCnoNEU)
text(0,6, paste("Correlation = ", round(cor(NEU, WBCnoNEU),3)), adj=0)
```
Fit model and evaluate VIFs:

```{r}
mod3 <- lm(CRP ~ NEU + WBCnoNEU)
VIF(mod3)

```

Evaluate the model:

```{r}
summary(mod3)
```

The standard errors for the coefficient of `NEU` is now considerably smaller, and the value of the coefficient is quite different from the model above. Still, the multiple $R^2$ is exactly equal to model `mod2`. The reason is that the coefficient of `NEU` now has a different meaning (expected difference in `CRP` corresponding to a difference in `NEU` of 1 G/L given constant concentrations of all other components of `WBC`). However, the coefficient of `WBCnoNEU` has the same meaning as in Model `mod2` (expected difference in `CRP` corresponding to a difference in `WBC` of 1 G/L given constant `NEU`), and hence its value and standard error in model `mod3` do not differ from those in `mod2`.
